---
title: "My data are censored!! What do I do??"
author: "Zane Billings"
date: "2023-10-30"
description: |
  I needed to make some censored data plots for a presentation, and it turned
  out to be easier to write the material as a blog post first and then convert
  that into a presentation afterwards.
license: "CC BY-SA"
draft: true
---

```{r setup, include = FALSE}
pth_base <- "2023-10-30-Censored-Data-Diagrams"
library(ggplot2)
# TODO add HAI breaks function to zlib?
ggplot2::theme_set(zlib::theme_ms())

library(lubridate)
study_start <- lubridate::as_date("2023-01-10")
study_end <- lubridate::as_date("2023-05-16")

pretty_date <- lubridate::stamp_date("October 07, 1998")
```

Recently, I've been working a lot on censored data in the context of antibody
measurements, which often have limits of detection or precision issues that
can be treated as censored data problems. I've been working with my
[awesome lab group](https://handelgroup.uga.edu/) on a larger project about
approaches for censored data analyses, but this blog post will cover the
very basics. I needed to make a presentation for a seminar about this topic,
and it turned out it was easier to write the blog post first and then
convert it into a presentation.

Here are the learning goals for this blog post, and the motivating issues for
me to write about. If I've done a decent job, by the end of this post, a
diligent reader should be able to:

1. Explain what censored data is, and give an example of how censoring might
occur;
1. List the types of censored data, and be able to identify these types in 
practical examples;
1. Write down a data generating model for censored data, and write down the
likelihood for the censored variables;
1. Explain how censored data are handled in survival, tobit-type, and Bayesian
models using the likelihood;
1. Identify pros and cons for frequentist and Bayesian approaches to handle
censored data.

## What is censored data?

**Censoring** is a type of data coarsening which occurs due to the way
data are observed. Such data are only partially known, i.e., we know a range
of values for a given data point, but not the exact value of that
data point. The most common example in the literature occurs in survival
analysis. Notably, when data are censored, we still have a record for a given
individual, with an imprecise measurement -- we will not consider **truncated**
data here, where the measurement process completely excludes measurements
which are not known precisely.

## Censored survival data

Survival analysis methods are used to analyze data where the outcome is the time
when a specific event occurs -- originally death, which is where the name comes
from. Survival analysis is also called reliability analysis in fields where
instruments or machines are the subject of consideration rather than people. In
epidemiology, these types of survival outcomes often come up in the context of a
**cohort study**. We recruit a bunch of people, follow them for a specific
amount of time, and observe when certain outcomes happen. Importantly, *if the
study ends before we observe an outcome for a given individual, that does not
mean the individual never gets an outcome.*

Instead, our observation of no outcome occurrence for that individual is a
feature of our observation process -- we would follow all individuals in our
study until they were no longer at risk for the event if we could, but real
world constraints limit the amount of follow-up or observation time we can
have. The outcome observation for this person is said to be **right censored**,
which is the classical type of censoring dealt with in epidemiology. The term
"right" censoring comes from the fact that this individual's time-to-event
is in the interval $(\text{end of study}, \ \infty)$, so the true value could
be anywhere to the right of our observed measurement on a number line.

Of course, a cohort study can be defined in calendar time or in event time, but
for the sake of simplicity we'll assume that our cohort is in calendar time.
Say for example, that we started our study (and had all patients recruited) on
`r pretty_date(study_start)` and completed all of our follow-up on
`r pretty_date(study_end)`.

```{r rc-data, include=FALSE}
set.seed(1254813)
rc <-
	tibble::tribble(
		~event_time,
		"2023-01-21",
		"2023-01-23",
		"2023-02-12",
		"2023-02-12",
		"2023-02-17",
		"2023-03-05",
		"2023-03-30",
		"2023-05-03",
		"2023-05-21",
		"2023-07-20"
	) |>
	dplyr::slice_sample(n = 10) |>
	dplyr::mutate(
		censored = factor(
			event_time >= study_end,
			levels = c(FALSE, TRUE),
			labels = c("Observed", "Censored")
		),
		event_time = lubridate::ymd(event_time),
		c_event_time = dplyr::if_else(
			event_time >= study_end,
			study_end,
			event_time
		),
		id = dplyr::row_number()
	)
```


```{r}
#| label: fig-rc
#| fig-cap: |
#|   Example study data which include right censored observations. For
#|   this example, we assume that no one experienced the event before the study
#|   began, so only right censoring is possible. Individuals who did not
#|   experience the event before the end of the study are right-censored, and
#|   their study times are colored red. Individuals who experienced the
#|   event before the end of the study are not censored, and their study
#|   times are colored black. The points (circles for completely observed
#|   data, x's for censored data) show the value we observe as part of the
#|   study.
ggplot(rc) +
	aes(x = study_start, color = censored, group = id) +
	geom_vline(xintercept = study_start, color = "gray", lwd = 2) +
	geom_vline(xintercept = study_end, color = "gray", lwd = 2) +
	geom_segment(
		aes(xend = event_time, y = id, yend = id),
		lwd = 1.5,
		show.legend = FALSE
	) +
	geom_point(
		aes(x = study_start, y = id),
		size = 3, stroke = 2, shape = 16,
		show.legend = FALSE
	) +
	geom_point(
		aes(x = c_event_time, y = id, shape = censored),
		size = 3, stroke = 2
	) +
	scale_color_manual(values = c("black", "red"), name = NULL) +
	scale_x_date(
		name = "Event time",
		expand = expansion(mult = 0.025),
		#date_breaks = "2 weeks",
		breaks = seq(study_start, study_end + 14, 14),
		date_labels = "%b %d"
	) +
	coord_cartesian(xlim = c(study_start, study_end + 14)) +
	scale_y_continuous(breaks = NULL, name = "Participant") +
	scale_shape_manual(values = c(16, 4), name = NULL)

ggsave(
	here::here("posts", pth_base, "rc.png"),
	width = 16, height = 9
)
```

Thinking about censored data in this way is useful for analyzing survival data,
but can be misleading when trying to understand the general idea of censored
data. Such a study would produce a distribution of survival times which is
**censored in the right tail.** With only 10 observations like we show in the
previous figure, visualizing the distribution of survival times might be
difficult. But if we draw survival times from some parametric model, and then
censor them, we can see this effect in practice. So suppose these are 10
individuals from a study that had 1000 individuals. The nature of the parametric
model and how I made this distribution is not important but is explained in
the callout box for anyone curious.

CALLOUT BOX FOR MODEL HERE

```{r rc-dist}
total_time <- lubridate::interval(study_start, study_end) / lubridate::days(1)
set.seed(12341234)
rc_vals <- lubridate::interval(study_start, rc$event_time) / lubridate::days(1)
est <- MASS::fitdistr(rc_vals, densfun = "weibull")
sim_dist_rc <- tibble::tibble(
	y_star = c(
		rc_vals,
		rweibull(990, shape = est$estimate[[1]], scale = est$estimate[[2]])
	),
	orig = c(rep(1, times = 10), rep(0, times = 990)),
	cens = ifelse(y_star <= total_time, 0, 1),
	y_cens = ifelse(cens, total_time, y_star)
)

dist_dat <- tibble::tibble(
	x = seq(0, 300, 0.01),
	y = dweibull(x, shape = est$estimate[[1]], scale = est$estimate[[2]])
)
ggplot(sim_dist_rc) +
	#geom_vline(xintercept = total_time) +
	geom_histogram(
		aes(x = y_cens, y = after_stat(density)), binwidth = 5,
		boundary = 0,
		color = "black", fill = "gray"
	) +
	geom_line(data = dist_dat, aes(x = x, y = y)) +
	labs(x = "Survival time")
```

Here, the histogram is showing the observed data from our censored sample.
There is a huge bin containing the censored values, which in this case are all
survival times that were larger than the amount of time each individual spent
in the study. The black curve shows the true density curve from which I
sampled the data, so we can see that the censored data are censored at a
specific measurement. Basically all of the density that would be above the
censoring point gets pushed into the closest bin just below the censoring point.
So this is how we can understand the term "right" censoring.

Similarly, 


The particular design where no participant has experienced the outcome when the
study begins is typically just called a cohort study, but we could specifically
call this an **incident cohort study**, because the design allows us to estimate
the incidence of the outcome in the source population.

In this type of incident cohort study, we cannot have **left censored data** by
definition. An individual with a left-censored time-to-event would have to be
at risk before the study begins.

In a prevalent cohort design, we
recruit individuals who have already experience the outcome of interest. I
was having trouble thinking of an example for when this study design might be
appropriate and there is no competing risk (which I want to avoid talking about
here because it makes things more complicated), but after some googling, I found
that one example might be assessment of the harms caused by an environmental or
occupational exposure.

<!-- EXAMPLE HERE
For example, suppose a cohort of factory workers during
a specific calendar time period are exposed to a dangerous chemical.
In the future, we begin to suspect that exposure to the chemical is associated
with developing a specific cancer later in life. There are likely sporadic
cases of this cancer which have been traced back to the chemical exposure, 
but we may want to enroll as many of the other exposed individuals as possible
into a surveillance program. So, any individuals who have already been diagnosed
with the specific cancer would be **left censored** in this study design: that
is, they experienced the outcome before we began the study.
-->

For now, the specific example is not too important but I'll try to think of
one and add it here later.

```{r lc-data, include=FALSE}
set.seed(1254814)
lc <-
	tibble::tribble(
		~t1, ~t2,
		"2022-11-16", "2023-05-29",
		"2023-03-27", "2023-05-02",
		"2023-01-04", "2023-03-07",
		"2022-12-29", "2023-02-15",
		"2023-01-21", "2023-04-17",
		"2023-01-11", "2023-05-05",
		"2023-02-04", "2023-03-30",
		"2023-02-01", "2023-05-03",
		"2023-03-21", "2023-05-21",
		"2023-05-01", "2023-07-20"
	) |>
	dplyr::slice_sample(n = 10) |>
	dplyr::mutate(
		r_censored = factor(
			t2 >= study_end,
			levels = c(FALSE, TRUE),
			labels = c("Observed", "Censored")
		),
		l_censored = factor(
			t1 <= study_start,
			levels = c(FALSE, TRUE),
			labels = c("Observed", "Censored")
		),
		censored = factor(
			(t2 >= study_end) | (t1 <= study_start),
			levels = c(FALSE, TRUE),
			labels = c("Observed", "Censored")
		),
		t1 = lubridate::ymd(t1),
		t2 = lubridate::ymd(t2),
		# # Deal with right censoring dot marks
		# c_event_time = dplyr::if_else(
		# 	t2 >= study_end,
		# 	study_end,
		# 	t2
		# ),
		# # Deal with left censoring dot marks
		# c_event_time = dplyr::if_else(
		# 	t2 <= study_start,
		# 	study_start,
		# 	t2
		# ),
		id = dplyr::row_number()
	)
```

```{r}
#| label: fig-lc
#| fig-cap: |
#|   The caption

ggplot(lc) +
	aes(x = t2, y = id, color = censored, group = id) +
	geom_vline(xintercept = study_start, color = "gray", lwd = 2) +
	geom_vline(xintercept = study_end, color = "gray", lwd = 2) +
	geom_segment(
		aes(xend = t1, y = id, yend = id),
		lwd = 1.5,
		show.legend = FALSE
	) +
	# geom_segment(
	# 	aes(
	# 		x = t1, xend = t1,
	# 		y = id - 0.5, yend = id + 0.5
	# 	),
	# 	lwd = 1.5,
	# 	show.legend = FALSE
	# ) +
	geom_label(
		aes(
			# Insane date math to get the graphical midpoint
			x = lubridate::as_date((
				pmax(as.numeric(t1), as.numeric(study_start - 14 * 2)) +
					pmin(as.numeric(t2), as.numeric(study_end + 14 * 1))
			) / 2),
			label = paste0("Participant #", id)
		)
	) +
	geom_point(
		aes(x = t2, y = id, shape = r_censored),
		size = 3, stroke = 2
	) +
	geom_point(
		aes(x = t1, y = id, shape = l_censored),
		size = 3, stroke = 2
	) +
	annotate(
		"label",
		x = study_start,
		y = 10.75,
		label = "Study start"
	) +
	annotate(
		"label",
		x = study_end,
		y = 10.75,
		label = "Study end"
	) +
	scale_color_manual(values = c("black", "red"), name = NULL) +
	scale_x_date(
		name = "Event time",
		expand = expansion(mult = 0.025),
		#date_breaks = "2 weeks",
		breaks = seq(study_start - 14 * 1, study_end + 14 * 1, 14),
		date_labels = "%b %d"
	) +
	coord_cartesian(xlim = c(study_start - 14 * 1, study_end + 14 * 1)) +
	scale_y_continuous(breaks = NULL, name = "Participant") +
	scale_shape_manual(values = c(16, 4), name = NULL)
```

## Censored non-survival data

```{r}

```

## An aside on the likelihood

## MLE on censored data

```{r}
set.seed(134123)
x_star <- rnorm(10000, 0, 1)
x <- ifelse(x_star < 0, 0, x_star)

star_dens <- density(x_star, kernel = "gaussian")
x_dens <- density(x, kernel = "gaussian", bw = star_dens$bw)

plot(
	NULL, NULL,
	xlim = c(min(star_dens$x, x_dens$x), max(star_dens$x, x_dens$x)),
	ylim = c(min(star_dens$y, x_dens$y), max(star_dens$y, x_dens$y)),
	xlab = "x", ylab = "density"
)
hist(x_star, breaks = seq(-5, 5, 0.1), freq = FALSE, add = TRUE,
	 col = "#E69F0060")
hist(x, breaks = seq(-5, 5, 0.1), freq = FALSE, add = TRUE,
	 col = "#56B4E960")
lines(star_dens, lty = 1, col = "orange", lwd = 2)
lines(x_dens, lty = 2, col = "blue", lwd = 2)
```

## The Bayesian alternative

## Conclusion

## Details {.appendix}

Last updated at `r Sys.time()`.

[source code](https://github.com/wzbillings/quarto-website/blob/main/posts/2023-10-30-Censored-Data-Diagrams)

```{r}
sessionInfo()
```


