---
title: "Multilevel models for a complex open cohort"
author: "Zane Billings"
date: last-modified
date-format: iso
description: |
  One of the data sources that I'm using for my cohort has a complicated
  multilevel structure. Since I had to figure out what to do with it, I
  thought I'd write about it in case my thoughts are useful for anyone dealing
  with similar data.
license: "CC BY-SA"
draft: true
---

```{r setup}
library(ggplot2)
ggplot2::theme_set(hgp::theme_ms())
```
I started this post by trying to write a StackExchange question. I thought it
would be fairly simple, but it kept getting longer and eventually I got stuck so
I thought I'd go back to the drawing board. Eventually I wrote down something
like this.

> If individuals didn't repeat, I think this would be a fairly simple structure
because it amounts to a normal 3-level nested model. The model should look
something like

And at that point I realized I didn't actually know how to write out that
model, so it was time to do some thinking and writing and actually figure out
what I was talking about. I've used multilevel models multiple times before,
and even took a class on longitudinal data analysis, but the complex structure
of this problem got me. So I'll start by writing out a (simplfied, generalized)
version of the model problem and exactly specifying my notation. Then we
can talk about how to write this model in both math-language and
`lme4`-language, and what I ultimately decided was the best model structure.

## Data description
> I am working with data from a study that has repeated measures at two scales,
and I'm not certain how to correctly specify the structure of this model. I have
a decent amount of experience with MLN/hierarchical models, but this one has
stumped me.
>
> The data were collected over multiple waves, and each wave has multiple
individuals. The design is a prospective open cohort, which means individuals
can repeat for any arbitrary pattern of waves. So, individual 1 might repeat
from wave 1 all the way to wave $J$ (the final year). Individual 27 might only
be present in wave 3. Individual 93 might be present in waves 1, 2, 3, 7, and 9,
etc. Any arbitrary combination is allowed. So this is one level of the repeated
measures structure of the data -- individuals are repeated over waves.
>
> Each individual also shows up twice during each season--there are multiple
time points in each year. So, each individual, for each year, is measured
between 2 and 5 times over the course of the year. The effect of repeated
measures over time, nested within each individual, needs to be treated
separately from the wave because there are qualitative differences between the
waves. (We know the dates when each wave was happening, and we have the dates
for each measurement, but we can't just combine things into one time variable.)
>
> Finally, at each time point, each individual also has multiple assays -- each
time point represents a serological sample, and multiple different (correlated,
comparable) assays are performed on the sample. So we have a cluster of
measurements nested within intra-wave time points, which are nested within
individuals.

# Simplifying models

One of the more useful things I learned in my math undergrad was
*Pólya's principle*:
> If you cannot solve the proposed problem, try to solve first some related problem. Could you imagine a more accessible related problem? --
> [George Pólya](https://en.wikipedia.org/wiki/How_to_Solve_It).

So instead of immediately tackling the large and complex model all at once,
we should first attempt to solve some smaller, related problems. Solving
these smaller problems will help us to understand how all the pieces can
fit together.

## Details {.appendix}

Last updated at `r Sys.time()`.

[source code](https://github.com/wzbillings/zlog/tree/master/_posts/posts/2024-01-15-Open-Cohort-MLM/index.Rmd)

```{r}
sessionInfo()
```
